Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-30 submission/manuscript.Rmd
add supplementary figure
remove notes

\clearpage
# Notes
## Things to do
- Move dazl to new figure if other staining works
- Use Ddx4 staining (with dazl?). If not should add RNA to diagram because its a key germline gene.
- Motif analysis
	- Discussion - talk about motifs

## Dazl
We were particularly interested in the aberrant transcription of _Dazl_, since it is essential for germ cell development and promotes the translation of germline mRNAs[@liDAZLMasterTranslational2019]. A significant portion of germline transcripts misexpressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein could promote the translation of other aberrant germline transcripts, influencing their ability to impact _Kdm5c_-KO cellular function. We thus tested DAZL protein expression in _Kdm5c_-KO EpiLCs through immunocytochemistry (Figure 3I). We observed about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein and it was localized to the cytoplasm (p = 0.0015, Welch's t-test), consistent with the pattern of DAZL expression in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest tissue-specific genes are misexpressed during _Kdm5c_-KO embryogenesis, including key drivers of germline identity that can be translated into protein.

- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020]. **note: maybe just put in results**. Could move around depending upon if I get pheno working.

## Discussion notes

- For other paper:
	- for methods: Heatmaps of gene expression were generated using the base R functions scale and hclust and visualized using the R package ComplexHeatmap (v2.12.1).
	- - Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654

- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/

- end with something like: this indicates kdm5c not only modulates pluripotency and self-renewal in ESCs, but also has a role in long-term silencing of germline genes 
	- then transition into the long term silencing mechanism paragraph

- KDM5C is dynamically regulated during the window of embryonic germline gene silencing, our EpiLC ChIP-seq is likely catching the tail end of KDM5C's main involvement.

- Our results indicate KDM5C is a pivotal repressor of germline programs in somatic cells. 
- Studies on other chromatin regulators have focused on key marker genes to identify germline gene misexpression, such as _Dazl_. 
- Using a germline-depleted mouse model, we curated a list of mouse germline-enriched genes to globally assess germline gene dysregulation. 
- Many of the genes dysregulated in the mature _Kdm5c_-KO brain were important for late stages of spermatogenesis, such as those important for sperm flagellar structure. Contrastingly, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germ cell identity, including _Dazl_ and _Stra8_.

- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation

- The including the demarcation between soma and germline fates.

	the _Kdm5c_-KO brain also 
	Previous work identified testis-enriched genes in _Kdm5c_ genes
	- 
	- However unlike the gonadal-biased DEGs,

- The demarcation of the germline versus soma is a key feature of multicellularity and sexually dimorphic reproduction

- Anything known about tissue-biased gene expression in other H3K4me regulators?
- Our data suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
- tissue-biased gene expression:

- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.

- Papers to read/reference:
	- Reconstitution of the Mouse Germ Cell Specification Pathway in Culture by Pluripotent Stem Cells: https://www.cell.com/fulltext/S0092-8674(11)00771-9
	- two cell gene list used by Suzuki et al Max paper is based on 2 cell sequencing: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3395470/

## Figure outline:

**Figure 1: Misexpression of tissue-specific genes in the _Kdm5c_-KO brain**
	* MA-plot and bar graphs of tissue-enriched genes
	* Example testis-specific genes (NCBI and bigwigs)
	* An example ovary tissue-specific gene 
	* An example muscle/liver tissue-specific gene (NCBI and bigwigs)

**Figure 2: The male _Kdm5c_-KO brain expresses male and female germline-enriched genes**
	* Gene ontology of testis DEGs in the amygdala and hippocampus - ontologies are germline ontologies
	* Expression of testis DEGs in germline-depleted testis (this is adult testis data)
	* scRNAseq of testis - # of testis DEGs that are germline-specific markers
	* Although far fewer, 5CKO brain also expresses ovary-enriched genes(NCBI and bigwigs of Zar1) 
	* These ovary enriched genes are also germline specific (NCBI/Li tissues are in adult ovary, so it would be best to show they're oocyte-specific in adult ovary. But I don't think there's a published adult female W/Wv dataset. Could try looking at scRNAseq or just do TPM in embryonic W/Wv data since oocytes are developed at this point? Or both?)
	* Defining what is/isn't a germline gene, and which are male/female biased using embryonic W/Wv data

**Figure 3: Kdm5c-KO epiblast-like cells express key drivers regulators of germline identity**
	* A) ESC to EpiLC differentiation Left - Morphology is unchanged, 
	* B) 5CKO EpiLCs express EpiLC differentiation genes similar to WT lvls
	* C) Male EpiLCs express germline genes (example Cyct again)
	* Overlap between brain and EpiLC germline genes - show they're mostly unique
	* GO of Brain and EpiLC germline genes (meiotic enriched) 
	* Bigwigs or TPM of master regulators
		* Show that while some are also 2-cell genes (Dazl), 2-cell specific genes aren't dysregulated (Zscan4). Important point because published KDM5C dazl paper is saying KDM5C is a 2-cell regulator, but as far as I can tell only genes shared between germline and 2-cell are dysregulated.  
	* Staining of Dazl (+ Stra8 if I can get it to work)

**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
	* Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
	* Increase in H3K4me2 in EpiLCs at germline genes 
	* Kdm5c binding in EpiLCs vs PNCs to show that germline repression is happening in early embryo
	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later). KDM5C RNA and protein ESC --> EpiLC (increasing then decreasing)
	* RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
	* DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)

**Figure 5: Ectopic, germline-like phenotypes in Kdm5c-KO ESCs/EpiLCs**
	* Sycp3 staining
	* DDX4 staining and repression of retrotransposons
	* Cilia??

Gaps in knowledge addressed: 
* Are other tissue-enriched genes dysregulated, or only testis, germline genes?
* Curating a robust list of male and female germline genes
	* Should talk about 2-cell genes vs germline genes - way to systematically categorize? 
* Mechanism behind long-term germline gene misexpression
	* Recent evidence suggests loss of KDM5C in ESCs express some germline genes
		* Unclear if catalytic activity is required for long-term silencing
		* Unclear if their dysregulation lasts throughout life or  the same between brain or not
	* When in development does it begin? - Recent evidence suggests some germline genes expressed in 5CKO ESCs but unclear if their dysregulation lasts throughout differentiation and if the identity of germline genes are different compared to the brain
* Are there functional consequences to germline gene misexpression?

Introduction:
* Chromatin regulators are important for cellular identity
	* H3K4me1-3 linked to active gene promoters and enhancers
	* Surprisingly, mutations in Chromatin regulators lead to many NDDs (including many H3K4 regulators)
		* Recent studies have shown some chromatin regulators are important for regulating neuron-specific gene expression/chromatin stat_compare_means 
	* However, loss of some chromatin regulators can also lead to ectopic expression of tissue-enriched genes
		* Very few studies have looked at these genes and it's unclear if these genes contribute to NDD impairments. 
	* Necessary to first characterize the mechanism behind their derepression to identify molecular footholds into testing their contribution to neuronal impairments and potential for therapeutic intervention

* Loss of KDM5C can result in the misexpression of genes typically only found in the testis
	* Misexpression of tissue-enriched genes hasn't been systematically characterized - Unclear if these genes are exceptions or if other tissue-specific genes are dysregulated
	* Interestingly, these genes (Cyct, D1pas1) typically function in the germline
	* Germ cells (meiotic cells) are typically distinguished from somatic cells very early on in embyrogenesis and is a key feature of multicellularity
	* Chromatin regulators are very important for decommissioning germline genes and act successively the embryo implants into the uterine wall
		* Most studies have focused on ESCs, which have a similar transcriptome to germ cells / 2-cells
		* recently, KDM5C was shown to repress DAZL in ESCs, independent of its catalytic activity
		* However, DNA methylation is lost in the mature 5CKO brain, DNA methylation is placed later and it's Unclear if it's required for long-term repression (maybe too specific, just trying to go into the fact that the mechanism is partially understood but unclear)
	* Systematic characterization of ectopic germline genes hasn't been done
		* unknown if other germline-enriched genes are dysregulated, including oocyte-specific genes
		* Crucially, it's unknown if misexpression of the germline program leads to functional consequences in 5CKO cells.

**Germline gene repression background:**

Interestingly, some of the ectopic testis transcripts identified in the _Kdm5c_-KO brain are typically expressed in germ cells[@scandagliaLossKdm5cCauses2017]. Unlike somatic cells, germ cells (e.g. sperm and eggs) undergo meiosis and pass on their genetic material to the next generation. The germline and the soma are typically distinguished during early embryogenesis, when germline genes are silenced in epiblast stem cells soon after implantation and only reactivated in a subset to form the germline. Chromatin regulators play a key role in decommissioning germline genes as the embryo transitions from na&#239;ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. KDM5C may also be involved in this early decommissioning of germline genes, as re-expression of KDM5C in neurons fails to surpress their dysregulation[@scandagliaLossKdm5cCauses2017]. In support of this, KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. In support of this, two independent screens in mouse embryonic stem cells (ESCs) recently identified KDM5C as a repressor of _Deleted in azoospermia like (Dazl)_, a key regulator of germline development. However, KDM5C's role in embryonic germline gene repression is currently unclear, given that Dazl is also expressed in ESCs and in the 2-cell stage and germline gene misexpression has yet to be globally characterized during _Kdm5c_-KO embryogenesis.

>>>>

\clearpage
